Parameter estimation

### 2023-08-06 SFE-article/Article_Model_Parameters.tex
06.08

\author[1]{Oliwer Sliczniuk*}[orcid=0000-0003-2593-5956]
\ead{oliwer.sliczniuk@aalto.fi}
\cormark[1]
\credit{a}

@@

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
%Given a system of partial differential equations, $F(t,x,\dot{x},p,u)=0$, where $x$ represents state variables, $p$ is the parameters, and $u$ are control variables, the process model is simultaneously solved for both $x_i$ and a set of sensitivity functions, $dx_i/dp_j$, overall times $t$.These sensitivity functions measure the influence of the parameter change on the model's output. As an example, the supercritical extraction process is presented. The impact of mass flow rate, pressure, and inlet temperature on the model's output is discussed. The sensitivity analysis results prove that the considered variables can affect the extraction's yield and be used as control variables in optimization problems. Moreover, the local sensitivity analysis results are analyzed from a phenomenological point of view to enhance understanding of the process model.
%Given a system of partial differential equations, $F(t,x,\dot{x},\Theta,u)=0$, where $x$ represents state variables, $\Theta$ are the parameters, and $u$ are control variables, we describe the supercritical extraction process. The process model describes a partially filled extractor with a fixed bed, which work under constant operating conditions. We assume that the flow is uniform across any cross-section, although the area available for the fluid phase can change along the extractor. We apply the concept of quasi-one-dimensional flow to mimic the modelling of a two-dimensional case.
%In this work, a distributed-parameter model, based on \citet{Reverchon1996} is used to describe a fluid-solid extraction process of caraway oil from caraway seeds with $CO_2$ as a solvent. The model parameters such as partition factor, internal diffusion coefficient, axial diffusion coefficient \textit{and saturation concentration} are obtained from parameter estimation. The parameters are estimated based on four experiments performed at $40^\circ C$/$200$ bar, $50^\circ C$/$200$ bar, $40^\circ C$/$300$ bar and $50^\circ C$/$300$ bar. Given yield data, the model-based parameter estimation uses the maximum likelihood estimation method under assumption of normal error.
This study aimed to investigate the supercritical extraction process of the caraway oil from caraway seeds from a partially filled extractor with a fixed bed operated under multiple constant operating conditions. The dataset collected from the experiments consists of four experiments conducted at different operating conditions: $40^\circ C$/$200$ bar, $50^\circ C$/$200$ bar, $40^\circ C$/$300$ bar, and $50^\circ C$/$300$ bar. A distributed-parameter model is used to describe the fluid-solid extraction process with $CO_2$ as a solvent. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. This model requires parameters such as partition factor, internal diffusion coefficient, axial diffusion coefficient, saturation concentration, and an initial state estimate. These parameters are estimated from the Peng-Robinson equation of state and by applying the maximum likelihood estimation method on yield data under the normal error assumption.

@@

where ${\color{orange}\rho_f}$ is the density of the fluid, ${\color{magenta}A_f}(z)$ is the function which describe change of the cross-section, $v$ is the velocity, ${\color{blue}P}$ is the total pressure, $e$ is the internal energy of the fluid, $t$ is time and $z$ is the spacial direction.

@@

\begin{figure}[!h]
	\centering
	\resizebox{0.95\columnwidth}{!}{%
		\begin{tikzpicture}[]
			\draw (0,2) -- (9,2);	% Top line
			\draw (0,0) -- (9,0); 	% Bottom line
			\draw (4.5,0) -- (4.5,2); 	% Bottom line
			\node at (4,1.5) {${\color{orange}\rho_f}$};
			\node at (5.5,1.5) {${\color{orange}\rho_f}+\delta{\color{orange}\rho_f}$};
			\node at (4,1.0) {$P$};
			\node at (5.5,1.0) {${\color{blue}P}+\delta {\color{blue}P}$};
			\node at (4,0.5) {$T$};
			\node at (5.5,0.5) {${\color{blue}T}+\delta {\color{blue}T}$};
			\node at (1,1.0) {$v \rightarrow$};
			\node at (8,1.0) {$v + \delta v \rightarrow$};
	\end{tikzpicture} }
	\caption{Small discontinuity in one-dimensional flow}
	\label{fig: Discontinuity_slow_flow}
\end{figure}

@@

This study presented basic aspects regarding the process parameters for obtaining caraway extracts using supercritical $CO_2$ at different operating conditions. A first-principle process model was developed based on governing equations and simplified using the Low-Mach number approximation. The values of unknown parameters in the process model were obtained by parameter estimation. The yield data from the experiments were compared against the yield data generated by the model. The IPOPT solver was used to solve the maximum likelihood estimation. The process model was found to have good agreement with experimental yield curves. Later, the parameters found for each experiment were combined to obtain more general relationships. The correlations for $D_i$, $\Upsilon$ and $D_e^M$ are presented as a function of $\rho$ or $R_e$. It should be noted that obtained correlations have been prepared with a limited amount of data. These correlations can be used to study the impact of different operating conditions on the extraction yield by different methods, such as sensitivity analysis. Moreover, the generalized model could be useful in finding the optimal operating conditions with respect to the process economy.

>>>>

\author[1]{Oliwer Sliczniuk}[orcid=0000-0003-2593-5956]
\ead{oliwer.sliczniuk@aalto.fi}
\cormark[1]
\credit{a}

@@

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
%Given a system of partial differential equations, $F(t,x,\dot{x},p,u)=0$, where $x$ represents state variables, $p$ is the parameters, and $u$ are control variables, the process model is simultaneously solved for both $x_i$ and a set of sensitivity functions, $dx_i/dp_j$, overall times $t$.These sensitivity functions measure the influence of the parameter change on the model's output. As an example, the supercritical extraction process is presented. The impact of mass flow rate, pressure, and inlet temperature on the model's output is discussed. The sensitivity analysis results prove that the considered variables can affect the extraction's yield and be used as control variables in optimization problems. Moreover, the local sensitivity analysis results are analyzed from a phenomenological point of view to enhance understanding of the process model.
%Given a system of partial differential equations, $F(t,x,\dot{x},\Theta,u)=0$, where $x$ represents state variables, $\Theta$ are the parameters, and $u$ are control variables, we describe the supercritical extraction process. The process model describes a partially filled extractor with a fixed bed, which work under constant operating conditions. We assume that the flow is uniform across any cross-section, although the area available for the fluid phase can change along the extractor. We apply the concept of quasi-one-dimensional flow to mimic the modelling of a two-dimensional case.
%In this work, a distributed-parameter model, based on \citet{Reverchon1996} is used to describe a fluid-solid extraction process of caraway oil from caraway seeds with $CO_2$ as a solvent. The model parameters such as partition factor, internal diffusion coefficient, axial diffusion coefficient \textit{and saturation concentration} are obtained from parameter estimation. The parameters are estimated based on four experiments performed at $40^\circ C$/$200$ bar, $50^\circ C$/$200$ bar, $40^\circ C$/$300$ bar and $50^\circ C$/$300$ bar. Given yield data, the model-based parameter estimation uses the maximum likelihood estimation method under assumption of normal error.
This study aimed to investigate the supercritical extraction process of the caraway oil from caraway seeds. The extraction was performed in a partially filled extractor with a fixed bed operated under multiple constant operating conditions. The dataset collected from four experiments consists of four experiments conducted at different operating conditions: $40^\circ C$/$200$ bar, $50^\circ C$/$200$ bar, $40^\circ C$/$300$ bar, and $50^\circ C$/$300$ bar. A distributed-parameter model is used to describe the fluid-solid extraction process with $CO_2$ as a solvent. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. This model requires parameters such as partition factor, internal diffusion coefficient, axial diffusion coefficient, saturation concentration, and an initial state estimate. These parameters are estimated from the Peng-Robinson equation of state and by applying the maximum likelihood estimation method on yield data under the normal error assumption.

@@

where ${\color{orange}\rho_f}$ is the density of the fluid, ${\color{magenta}A_f}(z)$ is the function which describe change of the cross-section, $v$ is the velocity, ${\color{blue}P}$ is the total pressure, ${\color{blue}e}$ is the internal energy of the fluid, $t$ is time and $z$ is the spacial direction.

@@

\begin{figure}[!h]
	\centering
	\resizebox{0.95\columnwidth}{!}{%
		\begin{tikzpicture}[]
			\draw (0,2) -- (9,2);	% Top line
			\draw (0,0) -- (9,0); 	% Bottom line
			\draw (4.5,0) -- (4.5,2); 	% Bottom line
			\node at (4,1.5) {${\color{orange}\rho_f}$};
			\node at (5.5,1.5) {${\color{orange}\rho_f}+\delta{\color{orange}\rho_f}$};
			\node at (4,1.0) {${\color{blue}P}$};
			\node at (5.5,1.0) {${\color{blue}P}+\delta {\color{blue}P}$};
			\node at (4,0.5) {${\color{blue}T}$};
			\node at (5.5,0.5) {${\color{blue}T}+\delta {\color{blue}T}$};
			\node at (1,1.0) {$v \rightarrow$};
			\node at (8,1.0) {$v + \delta v \rightarrow$};
	\end{tikzpicture} }
	\caption{Small discontinuity in one-dimensional flow}
	\label{fig: Discontinuity_slow_flow}
\end{figure}

@@

This study presented basic aspects regarding the laboratory work for obtaining caraway extracts using supercritical $CO_2$ at different operating conditions. A first-principle process model was developed based on governing equations and simplified using the Low-Mach number approximation. Based on the literature review, the extraction kinetic model was selected. The values of unknown parameters in the process model were obtained by parameter estimation. The yield data from the experiments were compared against the yield data generated by the model. The IPOPT solver was used to solve the maximum likelihood estimation. The process model was found to have good agreement with experimental yield curves. Later, the parameters found for each experiment were combined to obtain more general relationships. The correlations for ${\color{orange}D_i^R}$, ${\color{orange}\Upsilon}$ and ${\color{orange}D_e^M}$ are presented as a function of ${\color{orange}\rho}$ or ${\color{orange}R_e}$. It should be noted that obtained correlations have been prepared with a limited amount of data. These correlations can be used to study qualitatively the impact of different operating conditions on the extraction yield by different methods, such as sensitivity analysis. Moreover, the generalized model could be useful in finding the optimal operating conditions with respect to the process economy.

### 2023-08-06 SFE-article/Sections/Results.tex
06.08

    The parameter estimation problem was solved by fitting the process model to the dataset given in the Figure \ref{fig: estimation_results}. Each time-series present was fitted to the model separately. The blue dots indicate the data point obtained as a measurement from the laboratory experiments. The black curve represents the yield curve obtained from the initial guess of the parameters. The red curve correspond to yield curve obtained as a solution of the optimization problem. Every sub-figure is made of two figure of which one represent the cumulative yield, and the second one shows the derivative of the cumulative yield. As explained in the Chapter \ref{CH: Parameter_estimation}, the derivatives of the cumulative measurements are independent on the previous measurement and should be used for the parameter estimation.
    The parameters obtained from solving the optimization problem can be found in Table \ref{tab: Estimation_results}.

@@

	Values of the internal diffusion coefficients are distinguished for each experiment. The order of magnitude of ${\color{orange}D_i^R}$ obtained from the optimization is similar to values found by other researchers. \citet{Reverchon1996} performed the parameter estimation for the extraction process of sage oil from seeds and reported ${\color{magenta}D_i^R} \approx 6\times 10^{-13}[m^2/s]$. Figure \ref{fig: results_DI} represent the ${\color{orange}D_i^R}$ as the function of density. As the internal diffusion coefficient describes the diffusion inside of a particle, it is independent of external factors such as a fluid velocity around that particle. In such a case, the internal diffusion depends only on local conditions given by temperature and pressure, which defines the physical properties of a stagnant fluid inside of pores of a particle—considering that the internal diffusion can be represented as a function of the density of the fluid, which is uniquely defined at a given temperature and pressure in the one-phase region. The linear trend in Figure \ref{fig: results_DI} suggests that ${\color{orange}D_i^R}$ increases with fluid density.

	Similarly to the internal diffusion coefficient, the obtained values ${\color{orange}\Upsilon}$ can be represented as a function of the local operating conditions, or by uniquely defined physical properties such as density. Figure \ref{fig: results_upsilon} shows that of ${\color{orange}\Upsilon}$ grows with density. This can be explained if the solubility increases as the density grows. \citet{Shojaie2010}, showed that the solubility of carvone (one of the main components of caraway oil) rises together with pressure increase and decreases when temperature increases.

@@

		\begin{figure}[!h]
		\centering
		\begin{subfigure}[b]{0.49\columnwidth}
			\centering
			\includegraphics[trim = 10.5cm 14cm 10cm 1.5cm,clip,width=\columnwidth]{/Results_estimation/Trend_Lines_order_1.pdf}
			\caption{Dependency of ${\color{orange}D_i}$}
			\label{fig: results_DI}
		\end{subfigure}
		\begin{subfigure}[b]{0.49\columnwidth}
			\centering
			\includegraphics[trim = 3cm 8cm 17.7cm 7.7cm,clip,width=\columnwidth]{/Results_estimation/Trend_Lines_order_1.pdf}
			\caption{Dependency of ${\color{orange}\Upsilon}$}
			\label{fig: results_upsilon}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\columnwidth}
			\includegraphics[trim = 1.0cm 0cm 3cm 0cm, clip, width=\columnwidth]{/Results_estimation/D_i_Uppsilon_results.pdf}
			\caption{Dependency of ${\color{orange}D^R_i}$}
			\label{fig:Di_upsilon_res}
		\end{subfigure}
		\begin{subfigure}[b]{0.49\columnwidth}
			\centering
			\includegraphics[trim = 19.0cm 13.8cm 3.5cm 1.0cm,clip,width=\columnwidth]{/Results_estimation/Trend_Lines_D_e.pdf}
			\caption{Dependency of ${\color{orange}D^M_e}$}
			\label{fig: results_De}
		\end{subfigure}
		\caption{Results of parameter estimation}
		\label{fig: estimation_results_DI_GAMMA}
	\end{figure}

	The axial diffusion coefficient obtained from the optimizer seems is relatively high if compared to the internal diffusion coefficients. The values of ${\color{orange}D_e^M}$ have similar order of magnitude as reported in literature( \citet{ReisVasco2000}). %The relatively high values of the axial diffusion coefficients might be justified by taking into account low values of flow rate used in all the experiments. 
	The axial diffusion can be further analysed by defining the Pecelt number (${\color{orange}P_e} = v \times {\color{magenta}d_p} / {\color{orange}D_e^M} $) and the Reynolds number (${\color{orange}R_e} = v \times {\color{magenta}d_p} / {\color{orange}\mu} $), where ${\color{magenta}d_p}=0.15~[m]$ is the characteristic dimension defined to be the extractor diameter and $\mu$ is the fluid viscosity. %The obtained Pecelt numbers have the order of magnitude of $1 \times 10^{8}$, which suggest that diffusion is negligible, and advection dominates mass transport. 
	The obtained Pecelt numbers are $P_e \gg 1$, which suggest that mass diffusion is negligible, and advection dominates mass transport. An idea to consider a relationship between the Peclet (indirectly the mass diffusion coefficient) and Reynolds numbers has been proposed by \citet{Chung1968}. 
	The Figure \ref{fig: results_De} shows that the obtained values of the mass diffusion coefficient increase almost linearly with increment of the Reynolds number.

	As the total amount of the oil present in the system is unknown, and as such it has been obtained from the optimizer. Given the dataset and the process model, the optimizer found that the best fit is obtained if the total mass of the oil reach a lower bound equal to 80 g. The lower bound was estimated by round up of the biggest value of collected cumulative amount of the extraction product. As the same raw material was used in all experiments, the same initial amount of solute is assumed for all the cases.

	The inital state estimation resulted in values of $\tau$ in range from 0.69 to 0.76. These values are relatively similar to each other. It can be concluded that in the used experimental set-up around 30\% of the solute was extracted to the fluid phase during the preparation period.

	The noise present in each dataset is quantified by parameter $\sigma^2$. It can be observed dataset obtained at $40 ^\circ C$ / 200 bar and $30 ^\circ C$ / 300 bar have similar value of $\sigma^2\approx0.07$. The slightly better results were obtained for the experiment at $50 ^\circ C$ / 300 bar, when $\sigma^2\approx0.05$. The most noisy dataset corresponds to $50 ^\circ C$ / 200 bar, when $\sigma^2\approx0.12$. All values of $\sigma^2$ are relatively low, which suggest the mathematical model is capable to mimic the physical system up to the noise coming from the experiment.

>>>>

    The parameter estimation problem was solved by fitting the process model to the dataset given in the Figure \ref{fig: estimation_results}. Each time-series was fitted to the model separately. The blue dots indicate the data point obtained as a measurement from the laboratory experiments. The black curve represents the yield curve obtained from the initial guess of the parameters. The red curve correspond to yield curve obtained as a solution of the optimization problem. Every sub-figure is made of two figure of which one represent the cumulative yield, and the second one shows the derivative of the cumulative yield. As explained in Chapter \ref{CH: Parameter_estimation}, the derivatives of the cumulative measurements are independent on the previous measurement and should be used for the parameter estimation.
    The parameters obtained from solving the optimization problem can be found in Table \ref{tab: Estimation_results}.

@@

	Values of the reference internal diffusion coefficients are distinguished for each experiment. The order of magnitude of ${\color{orange}D_i^R}$ obtained from the optimization is similar to values found by other researchers. \citet{Reverchon1996} performed the parameter estimation for the extraction process of sage oil from seeds and reported ${\color{magenta}D_i^R} \approx 6\times 10^{-13}[m^2/s]$. Figure \ref{fig: results_DI} represent the ${\color{orange}D_i^R}$ as the function of density. As the reference internal diffusion coefficient describes the diffusion inside of a particle, it is independent of external factors such as a fluid velocity around that particle. The reference internal diffusion depends only on local conditions given by temperature and pressure, which defines the physical properties of a stagnant fluid inside of pores of a particle—considering that the reference internal diffusion can be represented as a function of the density of the fluid, which is uniquely defined at a given temperature and pressure in the one-phase region. The linear trend in Figure \ref{fig: results_DI} suggests that ${\color{orange}D_i^R}$ increases with fluid density.

	Similarly to the internal diffusion coefficient, the obtained values of ${\color{orange}\Upsilon}$ depends only on temperature and pressure, or by uniquely defined physical properties such as density. Figure \ref{fig: results_upsilon} shows that of ${\color{orange}\Upsilon}$ grows with density. This can be explained if the solubility increases as the density grows. \citet{Shojaie2010}, showed that the solubility of carvone (one of the main components of caraway oil) rises together with pressure increase and decreases when temperature increases.

@@

		\begin{figure}[!h]
		\centering
		\begin{subfigure}[b]{0.49\columnwidth}
			\centering
			\includegraphics[trim = 11.2cm 13.7cm 11cm 1.0cm,clip,width=\columnwidth]{/Results_estimation/Trend_Lines_order_1.pdf}
			\caption{Dependency of ${\color{orange}D_i}$}
			\label{fig: results_DI}
		\end{subfigure}
		\begin{subfigure}[b]{0.49\columnwidth}
			\centering
			\includegraphics[trim = 3.5cm 7.7cm 18.7cm 7.3cm,clip,width=\columnwidth]{/Results_estimation/Trend_Lines_order_1.pdf}
			\caption{Dependency of ${\color{orange}\Upsilon}$}
			\label{fig: results_upsilon}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\columnwidth}
			\includegraphics[trim = 1.0cm 0cm 3cm 0cm, clip, width=\columnwidth]{/Results_estimation/D_i_Uppsilon_results.pdf}
			\caption{Dependency of ${\color{orange}D^R_i}$}
			\label{fig:Di_upsilon_res}
		\end{subfigure}
		\begin{subfigure}[b]{0.49\columnwidth}
			\centering
			\includegraphics[trim = 19.0cm 13.8cm 3.5cm 1.0cm,clip,width=\columnwidth]{/Results_estimation/Trend_Lines_D_e.pdf}
			\caption{Dependency of ${\color{orange}D^M_e}$}
			\label{fig: results_De}
		\end{subfigure}
		\caption{Results of parameter estimation}
		\label{fig: estimation_results_DI_GAMMA}
	\end{figure}

	The axial diffusion coefficient obtained from the optimizer is relatively high compared to the internal diffusion coefficients. The values of ${\color{orange}D_e^M}$ have a similar order of magnitude as reported in the literature (\citet{ReisVasco2000}). %The relatively high values of the axial diffusion coefficients might be justified by taking into account the low values of the flow rate used in all the experiments. 
	The axial diffusion can be further analyzed by defining the Pecelt number (${\color{orange}P_e} = v \times {\color{magenta}d_p} / {\color{orange}D_e^M} $) and the Reynolds number (${\color{orange}R_e} = v \times {\color{magenta}d_p} / {\color{orange}\mu} $), where ${\color{magenta}d_p}=0.15~[m]$ is the characteristic dimension defined to be the extractor diameter and $\mu$ is the fluid viscosity. %The obtained Pecelt numbers have the order of magnitude of $1 \times 10^{8}$, which suggests that diffusion is negligible and advection dominates mass transport. 
	The obtained Pecelt numbers are ${\color{orange}P_e} \gg 1$, which suggests that mass diffusion is negligible and advection dominates mass transport. This observation suggests that the process model could be further simplified by neglecting the diffusion term. Nevertheless, such a simplification has not been introduced in this work.
	\citet{Chung1968} presented an idea for a relationship between the Peclet (indirectly the mass diffusion coefficient) and Reynolds numbers. By modifying the idea of \citet{Chung1968}, a direct relation between the diffusion coefficient and the Reynolds number is used in the work as shown in Figure \ref{fig: results_De}. The obtained values of the mass diffusion coefficient increase almost linearly with the increment of the Reynolds number, which is in agreement with the observation that the advection dominates the diffusion.

	As the total amount of oil present in the system is unknown, and as such, it has been obtained from the optimizer. Given the dataset and the process model, the optimizer found that the best fit is obtained if the total mass of the oil reaches a lower bound equal to 80 g. The lower bound was estimated by rounding up the biggest value of the collected cumulative amount of the extraction product. As the same raw material was used in all experiments, the same initial amount of solute was assumed for all the cases.

	The initial state estimation resulted in values of $\tau$ in the range from 0.69 to 0.76. These values are relatively similar to each other. In the equipment used in this experiment, around 30\% of the solute was extracted to the fluid phase during the preparation period.

	The noise present in each dataset is quantified by the parameter $\sigma^2$. It can be observed dataset obtained at $40 ^\circ C$ / 200 bar and $30 ^\circ C$ / 300 bar have similar value of $\sigma^2\approx0.07$. The slightly better results were obtained for the experiment at $50 ^\circ C$ / 300 bar when $\sigma^2\approx0.05$. The noisiest dataset corresponds to $50 ^\circ C$ / 200 bar when $\sigma^2\approx0.12$. All values of $\sigma^2$ are relatively low, suggesting the mathematical model can mimic the physical system up to the noise coming from the experiment.
